\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-20 paper/main.tex
edits through section 2

\documentclass[preprint,12pt,a4paper]{elsarticle}
\usepackage{amssymb}
\usepackage{lineno}
\usepackage{float}
\restylefloat{table}
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage[htt]{hyphenat}
\usepackage{placeins}
\newcommand{\comment}[1]{}
\usepackage{geometry}
\geometry{letterpaper,left=1in,right=1in,top=1in,bottom=1in,headsep=0pt,footskip=0pt}

@@

Within the Python~\cite{vanR95} community, external packages that are
published in the most popular repositories~\cite{Pyth03, cond15} are
associated with version numbers and tags that allow users to guarantee
they are installing exactly the same code across different computing
environments~\cite{CoghStuf13}.  While it is \textit{possible} to
manually install the intended version of every dependency of a Python
script or package, manually tracking down those dependencies can
impose a substantial burden on the user and creates room for mistakes
and inconsistencies. Further, when dependency versions are left
unspecified, replicating the original computing environment becomes
difficult or impossible.

@@

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the default notebook experience. Like standard
Jupyter notebooks, \texttt{davos}-enhanced notebooks allow researchers
to include text, executable code, and media within a single file. No
further setup or installation is required, beyond what is needed to
run standard Jupyter notebooks. And like virtual environments,
\texttt{davos} provides a convenient mechanism for fully specifying
(and installing, as needed) a complete set of Python dependencies,
including package versions.

@@

The \texttt{davos} package is named after Davos Seaworth, a smuggler
referred to as ``the Onion Knight" from the series \textit{A Song of
  Ice and Fire} by George R. R. Martin~\cite{Mart98}.  The
\texttt{smuggle} keyword-like operator implemented in \texttt{davos}
is a play on Python's \texttt{import} keyword: whereas importing can
bring a package into the Python workspace within the existing rules
and frameworks provided by the Python language, ``smuggling'' provides
an alternative that expands the scope and reach of ``importing.''
Like the character Davos Seaworth (who became famous for smuggling
onions through a blockade on his homeland), we use ``onion'' comments
to precisely control how packages are smuggled into the Python
workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The \texttt{davos} package (orange)
  comprises two interdependent subpackages (purple).  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring \texttt{davos}'s behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The \texttt{davos} package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement\comment{provide?} the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Section~\ref{subsec:smuggle}) and onion comment (see
Section~\ref{subsec:onion}), and a runtime interface for configuring
\texttt{davos}'s behavior (see Section~\ref{subsec:config}). However,
certain critical\comment{other important} aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
\texttt{davos} is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent parts of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, \texttt{davos} detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend \texttt{davos} to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment may be
used to ensure that \texttt{davos} installs the proper package if it
cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will download, install, and load a suitable
version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this design provides the user with additional control over which
packages are imported. For example, each of the
\mbox{\texttt{-I}/\texttt{--ignore-installed}},
\texttt{-U}/\texttt{--upgrade}, and \texttt{--force-reinstall} flags
will cause \texttt{davos} to skip searching for a smuggled package
locally before installing a new copy:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, the \texttt{--no-input} flag will temporarily enable
\texttt{davos}'s non-interactive mode (see
Section~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

The \texttt{davos} config object provides a high-level interface for
controlling various aspects of \texttt{davos}'s behavior. After
importing \texttt{davos}, the \texttt{davos.con\-fig} object (a
singleton) exposes configurable options as attributes that can be
modified, checked at runtime, or displayed (see
Sec.~\ref{sec:illustrative-example} for an illustrative example or
Sec.~\ref{subsec:implementation} for implementation details and
additional information). These include:

@@

\item \texttt{.auto\_rerun}: This attribute controls how
  \texttt{davos} behaves when attempting to \texttt{smuggle} a new
  version of a package that was previously imported and cannot be
  reloaded. This can happen if the package includes extension modules
  that dynamically link C or C++ objects to the Python interpreter,
  and if the code that generates those objects was changed between the
  previously imported and to-be-smuggled versions.  If this attribute
  is set to \texttt{True}, \texttt{davos} will automatically restart
  the notebook kernel and rerun all code up to (and including) the
  current \texttt{smuggle} statement. If \texttt{False} (the default),
  \texttt{davos} will instead issue a warning, pause execution, and
  prompt the user to either restart and rerun the notebook, or
  continue running with the previously imported package version until
  the next time the kernel is restarted manually.  Note that, as of
  this writing, the \texttt{.auto\_rerun} attribute is not supported
  in Google Colaboratory notebooks.

@@

\item \texttt{.suppress\_stdout}: If this attribute is set to
  \texttt{True} (default: \texttt{False}), \texttt{davos} suppresses
  printed (console) outputs from itself and the installer program.
  This can be useful when smuggling packages that need to install many
  dependencies and/or generate extensive output. However, if the installer
  program throws an error, both the stdout and stderr streams will still be
  displayed along with a stack trace.
\end{itemize}

@@

Although \texttt{davos} is designed to \textit{appear} to add a new
Python keyword to a notebook's vocabulary, this illusion is actually a
consequence of several ``hacks'' that make use of the IPython backend
for processing and executing notebook code.  Specifically, when
\texttt{davos} is first imported, or when it is activated after being
set to an inactive state, two actions are triggered.  First, the
\texttt{smuggle()} function is injected into the IPython user
namespace.  Second, the \texttt{davos} parser is registered as a
custom IPython input transformer.

>>>>

\documentclass[preprint,12pt,a4paper]{elsarticle}
\usepackage{amssymb}
\usepackage{lineno}
\usepackage{float}
\restylefloat{table}
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage{placeins}
\usepackage{geometry}
\geometry{left=1in,right=1in,top=1in,bottom=1in,headsep=0pt}

@@

Within the Python~\cite{vanR95} community, external packages that are
published in the most popular repositories~\cite{Pyth03, cond15} are
associated with version numbers and tags that allow users to guarantee
they are installing exactly the same code across different computing
environments~\cite{CoghStuf13}.  While it is \textit{possible} to
manually install the intended version of every dependency of a Python
script or package, manually tracking down those dependencies can
impose a substantial burden on the user and create room for mistakes
and inconsistencies. Further, when dependency versions are left
unspecified, replicating the original computing environment becomes
difficult or impossible.

@@

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the default notebook experience. Like standard
Jupyter notebooks, \texttt{davos}-enhanced notebooks allow researchers
to include text, executable code, and media within a single file. No
further setup or installation is required, beyond what is needed to
run standard Jupyter notebooks. And like virtual environments,
\texttt{davos} provides a convenient mechanism for fully specifying
(and installing, as needed) a complete set of Python dependencies,
including specific package versions.

@@

The \texttt{davos} package is named after Davos Seaworth, a smuggler
referred to as ``the Onion Knight" from the series \textit{A Song of
  Ice and Fire} by George R. R. Martin~\cite{Mart98}.  The
\texttt{smuggle} keyword provided by \texttt{davos}
is a play on Python's \texttt{import} keyword: whereas importing can
load a package into the Python workspace within the existing rules
and frameworks provided by the Python language, ``smuggling'' provides
an alternative that expands the scope and reach of ``importing.''
Like the character Davos Seaworth (who became famous for smuggling
onions through a blockade on his homeland), we use ``onion'' comments
to precisely control how packages are smuggled into the Python
workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The \texttt{davos} package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring \texttt{davos}'s behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The \texttt{davos} package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Section~\ref{subsec:smuggle}) and onion comment (see
Section~\ref{subsec:onion}), and a runtime interface for configuring
\texttt{davos}'s behavior (see Section~\ref{subsec:config}). However,
certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
\texttt{davos} is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent parts of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, \texttt{davos} detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend \texttt{davos} to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment may be
used to ensure that \texttt{davos} installs the proper package if it
cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will download, install, and load a suitable
version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this enables \texttt{davos} to adapt its behavior based on how particular
flags will affect the behavior of the installer program. For example, if an
onion comment contains either the \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall} flag,
\texttt{davos} will not bother checking for a local copy of the smuggled
package before installing a new one:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, the \texttt{--no-input} flag will temporarily enable
\texttt{davos}'s non-interactive mode (see
Section~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

The \texttt{davos} config object provides a high-level interface for
controlling various aspects of \texttt{davos}'s behavior. After
importing \texttt{davos}, the \texttt{davos.con\-fig} object (a
singleton) exposes configurable options as attributes that can be
modified, displayed in the notebook, or checked programmatically at runtime
(see Sec.~\ref{sec:illustrative-example} for an illustrative example or
Sec.~\ref{subsec:implementation} for implementation details and
additional information). These include:

@@

\item \texttt{.auto\_rerun}: This attribute controls how
  \texttt{davos} behaves when attempting to \texttt{smuggle} a new
  version of a package that was previously imported and cannot be
  reloaded. This can happen if the package includes extension modules
  that dynamically link C or C++ objects to the Python interpreter,
  and the code that generates those objects was changed between the
  previously imported and to-be-smuggled versions.  If this attribute
  is set to \texttt{True}, \texttt{davos} will automatically restart
  the notebook kernel and rerun all code up to (and including) the
  current \texttt{smuggle} statement. If \texttt{False} (the default),
  \texttt{davos} will instead issue a warning, pause execution, and
  prompt the user to either restart and rerun the notebook, or
  continue running with the previously imported package version until
  the next time the kernel is restarted manually.  Note that, as of
  this writing, the \texttt{.auto\_rerun} attribute is not supported
  in Google Colaboratory notebooks.

@@

\item \texttt{.suppress\_stdout}: If this attribute is set to
  \texttt{True} (default: \texttt{False}), \texttt{davos} suppresses
  printed (console) outputs from both itself and the installer program.
  This can be useful when smuggling packages that need to install many
  dependencies and/or generate extensive output. However, if the installer
  program throws an error, both its stdout and stderr streams will be
  displayed alongside the Python traceback to allow for debugging.
\end{itemize}

@@

Although \texttt{davos} is designed to \textit{appear} to add a new
keyword to Python's vocabulary, this illusion is actually created through
several ``hacks'' that make use of the notebook's IPython backend
for processing and executing users' code.  Specifically, when
\texttt{davos} is first imported, or when it is activated after having been
set to an inactive state, two actions are triggered:  First, the
\texttt{smuggle()} function is injected into the IPython user
namespace.  Second, the \texttt{davos} parser is registered as a
custom IPython input transformer.
